Implementation of a lattice Boltzmann method for immiscible two-phase flow simulations using the level set method

### 2015-11-15 Paper/paper.tex
noch ein paar Kleinigkeiten

In this publication we use a hybrid lattice Boltzmann level set method. The motion of the interface is modeled by the level set method. The input data for the level set method is given by the lattice Boltzmann method (LBM), which is applied independently for each fluid. This method has the advantage that it allows for large ranges of viscosity- and density-ratios between the fluid phases, while describing a sharp interface.

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ of the Boltzmann equation in phase space, with $(\vec{x},\vec{v})$ as the phase space variables over time. For the LBM algorithm the geometrical domain is discretized with a cartesian grid and the velocity space with a restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (\cref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=4cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM.}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix,
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
  \label{disc_vel}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the moments of the particle distribution functions with,
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=1}^{9} f_i(\vec{x},t) \text{, } \quad
	\vec{u} = \sum_{i=1}^{9} f_i(\vec{x},t)\vec{c_i} \text{,}
\end{equation} 
while the pressure is related to the lattice density $\rho$ as $p=\varrho * \rho/3$.

@@

Let $\Gamma_t$ be our interface at time t. Within the level set method, $\Gamma_t$ is the zero isosurface of the level set function $\varphi$. Thus, for $\vec x \in \Gamma_t$ we have $\varphi(\vec x,t) = 0$, in direction of the surface normal we get $\varphi > 0$, and in the other direction we get $\varphi < 0$. For this level set function we get the level set equation introduced by Osher and Sethian in \cite{OsherSethian} for the evolution of surfaces,
\begin{equation}
  \frac{\partial \varphi}{\partial t} + \vec v \cdot \nabla \varphi = 0 \text{,}
\end{equation}
with the velocity $\vec v = \vec u(\vec x(t))$ for $\vec x(t) \in \Gamma_t$, which we obtain from the LBM. It is easily visible, that the normal of the surface is determined as $\vec n(x) = \frac{\nabla \varphi(x)}{\|\varphi(x)\|}$ and accordingly the curvature as the divergence of the normal $\kappa = \nabla \cdot \nabla \varphi = \Delta \varphi$.

@@

\subsection{Couette channel flow}
\begin{figure}[b!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette1.pdf}\hspace*{\fill}
  \caption{Two horizontal fluid layers $\Omega_i$ separated by an interface. The top boundary is moving horizontally, while the lower one is at rest}
	\label{couette1}
\end{figure}
The Couette channel flow consists of a lower wall $y=0$ at rest and an upper wall $y=1$ that is moving tangentially with a prescribed constant velocity. At these walls no-slip boundary conditions hold. Along the $x$-direction periodic boundary conditions are prescribed. The two fluids layers are oriented such that their interface is parallel to the no-slip walls, and no external pressure gradient is applied.
For this flow scenario an analytic solution can be derived from the Navier-Stokes equations, namely they simplify to $\frac{\partial }{\partial y}\left( \eta\frac{\partial u}{\partial y}\right) =0$. In combination with the no-slip conditions at top and bottom and wall this means, that the horizontal velocity has a linear profile with a kink at the fluid interface. The respective slopes between both phases are related by $\frac{\partial_y u^{(1)}}{\partial_y u^{(2)}} = \frac{\eta^{(2)}}{\eta^{(1)}}$.

@@

Figure \ref{couette2} shows a second scenario with $5 \times 10$ cells along $x \times y$ direction and viscosities $\mu^{(1)} = 1$ and $\mu^{(2)} = 0.2$ with the interface located at $y=0.75$. Again the numerically approximated values agree very well with the analytic ones and the simulation converged at a relative error of $1.4\cdot10^{-4}$ after 2000 LBM iterations.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette2.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ in a Coutte flow}
	\label{couette2}
\end{figure}
We conclude from this that the implemented LBM boundary conditions successfully model viscosity jumps at the fluid interface. Furthermore, we were able to use viscosity ratios greater than $100:1$ in our tests.
\subsection{Young-Laplace experiment for a bubble}
The Couette flow scenario does not account for the influence of surface tension as the fluid interface is not curved. If it would get distorted, surface tension would have a stabilizing effect although. 
Hence, we use the Young-Laplace law for the pressure drop across curved surfaces, to validate the implementation of surface tension. Due to surface tension the pressure inside of convex surfaces is generally higher than on the outside and the pressure drop $\Delta p$ of a bubble of radius $r$ is proportional to the surface tension coefficient $\sigma$ according to the following relationship,
\begin{equation}
	\Delta p = \frac{2\sigma}{r} = 2 \sigma \kappa .
\end{equation}

Figure \ref{laplace} shows the simulation results we obtained for a $40 \times 40$ cells grid over a $[0,1]\times[0,1]$-domain with periodic boundaries. At the center, a sphere of radius $0.25$ was placed with $\varrho_2 = 1.1$ while the exterior contained a fluid with $\varrho_1 = 1$. It is visible, how the approximated values attain the analytical solution in a oscillatory motion decaying amplitude, until both values agree. The algorithm needed 1000 LBM-iterations, where every ten iterations, a call to the level-set routine was executed. Both fluids had an identical viscosity of $\nu = 1/6$.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 7.75cm 0mm 8.06cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{laplace.pdf}\hspace*{\fill}
	\caption{The pressure drop of a spherical bubble over LBM iterations}
	\label{laplace}
\end{figure}
We noted that the velocity close to the interface contained erroneous components, which lead to a slight distortion of the spherical shape. We explain those fluxes as inaccuracies stemming from the used level set subroutines, according to the analysis shown in \cite{Thoemmes2}.
Additionally, we want to state, that we were not able to attain stable results for surface tension coefficients $\sigma > \mathrm{O(10^{-3})}$. Large densitiy ratios were limited to big viscosity regimes, or refined spatial resolution in our tests.

>>>>

In this publication we use a hybrid lattice Boltzmann level set method. The motion of the interface is modeled by the level set method. The input data for the level set method is given by the lattice Boltzmann method (LBM), which is applied independently for each fluid. This method has the advantage that it allows large ranges of viscosity- and density-ratios between the fluid phases, while describing a sharp interface.

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ of the Boltzmann equation in phase space, with $(\vec{x},\vec{v})$ as the phase space variables over time. For the LBM algorithm the geometrical domain is discretized with a cartesian grid and the velocity space with a restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (\cref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=4cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM.}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix,
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
  \label{disc_vel}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the moments of the particle distribution functions with,
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=1}^{9} f_i(\vec{x},t) \text{, } \quad
	\vec{u} = \sum_{i=1}^{9} f_i(\vec{x},t)\vec{c_i} \text{,}
\end{equation} 
while the pressure is related to the lattice density $\rho$ as $p=\varrho \cdot \rho/3$.

@@

Let $\Gamma_t$ be our interface at time t. Within the level set method, $\Gamma_t$ is the zero isosurface of the level set function $\varphi$. Thus, for $\vec x \in \Gamma_t$ we have $\varphi(\vec x,t) = 0$, in direction of the surface normal we get $\varphi > 0$, and in the other direction we get $\varphi < 0$. For this level set function we get the level set equation introduced by Osher and Sethian in \cite{OsherSethian} for the evolution of surfaces,
\begin{equation}
  \frac{\partial \varphi}{\partial t} + \vec v \cdot \nabla \varphi = 0 \text{,}
\end{equation}
with the velocity $\vec v = \vec u(\vec x(t))$ for $\vec x(t) \in \Gamma_t$, which we obtain from the LBM. It is easily visible that the normal of the surface is determined as $\vec n(x) = \frac{\nabla \varphi(x)}{\|\varphi(x)\|}$ and accordingly the curvature as the divergence of the normal $\kappa = \nabla \cdot \nabla \varphi = \Delta \varphi$.

@@

\subsection{Couette channel flow}
\begin{figure}[b!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette1.pdf}\hspace*{\fill}
  \caption{Two horizontal fluid layers $\Omega_i$ separated by an interface. The top boundary is moving horizontally, while the lower one is at rest}
	\label{couette1}
\end{figure}
The Couette channel flow consists of a lower wall $y=0$ at rest and an upper wall $y=1$ that is moving tangentially with a prescribed constant velocity. At these walls no-slip boundary conditions hold. Along the $x$-direction periodic boundary conditions are prescribed. The two fluids layers are oriented such that their interface is parallel to the no-slip walls, and no external pressure gradient is applied.
For this flow scenario an analytic solution can be derived from the Navier-Stokes equations, namely they simplify to $\frac{\partial }{\partial y}\left( \eta\frac{\partial u}{\partial y}\right) =0$. In combination with the no-slip conditions at top and bottom wall this means that the horizontal velocity has a linear profile with a kink at the fluid interface. The respective slopes between both phases are related by $\frac{\partial_y u^{(1)}}{\partial_y u^{(2)}} = \frac{\eta^{(2)}}{\eta^{(1)}}$.

@@

Figure \ref{couette2} shows a second scenario with $5 \times 10$ cells along $x \times y$ direction and viscosities $\mu^{(1)} = 1$ and $\mu^{(2)} = 0.2$ with the interface located at $y=0.75$. Again the numerically approximated values agree very well with the analytic ones and the simulation converged at a relative error of $1.4\cdot10^{-4}$ after 2000 LBM iterations.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette2.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ in a Coutte flow}
	\label{couette2}
\end{figure}
We conclude from this that the implemented LBM boundary conditions successfully model viscosity jumps at the fluid interface. Furthermore, we were able to use viscosity ratios greater than $100:1$ in our tests.
\subsection{Young-Laplace experiment for a bubble}
The Couette flow scenario does not account for the influence of surface tension as the fluid interface is not curved. If it would get distorted, surface tension would have a stabilizing effect although. 
Hence, we use the Young-Laplace law for the pressure drop across curved surfaces to validate the implementation of surface tension. Due to surface tension the pressure inside of convex surfaces is generally higher than on the outside and the pressure drop $\Delta p$ of a bubble of radius $r$ is proportional to the surface tension coefficient $\sigma$ according to the following relationship,
\begin{equation}
	\Delta p = \frac{2\sigma}{r} = 2 \sigma \kappa .
\end{equation}

Figure \ref{laplace} shows the simulation results we obtained for a $40 \times 40$ cells grid over a $[0,1]\times[0,1]$-domain with periodic boundaries. At the center, a sphere of radius $0.25$ was placed with $\varrho_2 = 1.1$ while the exterior contained a fluid with $\varrho_1 = 1$. It is visible, how the approximated values attain the analytical solution in a oscillatory motion decaying amplitude, until both values agree. The algorithm needed 1000 LBM-iterations, where every ten iterations a call to the level-set routine was executed. Both fluids had an identical viscosity of $\nu = 1/6$.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 7.75cm 0mm 8.06cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{laplace.pdf}\hspace*{\fill}
	\caption{The pressure drop of a spherical bubble over LBM iterations}
	\label{laplace}
\end{figure}
We noted that the velocity close to the interface contained erroneous components, which lead to a slight distortion of the spherical shape. We explain those fluxes as inaccuracies stemming from the used level set subroutines, according to the analysis shown in \cite{Thoemmes2}.
Additionally, we want to state that we were not able to attain stable results for surface tension coefficients $\sigma > \mathrm{O(10^{-3})}$. Large densitiy ratios were limited to big viscosity regimes or refined spatial resolution in our tests.
